---
title: "Language R"
subtitle: Assignment
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introdução

The Hungarian physician [Dr. Ignaz Semmelweis](http://www.incor.usp.br/conteudo-medico/decourt/semmelweis.html) was born in 1818 and worked at the Vienna General Hospital. His work on puerperal fever was of great importance. Puerperal fever is a potentially fatal disease that affects women after childbirth. In the early 1840s, around 10% of women who had given birth at the Vienna General Hospital died from this disease. Dr. Semmelweis discovered the cause of puerperal fever: contaminated hands of doctors who attended women during childbirth. However, his discovery was not accepted for many years.

## Observing the situation

Let's reanalyze the data that led Dr. Semmelweis to discover the importance of handwashing.

```{r,message=FALSE}

# Loading the tidyverse package
library(tidyverse)

# Reading the dataset about annual deaths and saving it in the data_death object
data_death <- read_csv("mortes_anuais_clinica.csv")

# Renaming the columns
data_death <- data_death %>%
  rename(
    year = ano,
    births = nascimentos,
    death = mortes,
    clinical = clinica
  )

#Renaming some values 
data_death <- data_death %>%
  mutate(
    clinical = recode(
      clinical,
      "clinica 1" = "clinic 1",
      "clinica 2" = "Clinic 2"
    )
  )
# Printing the dataset
data_death
```

The dataset `data_death` shows the number of childbirths in two clinics at the Vienna General Hospital between 1841 and 1846. We can observe a high number of women who died during childbirth, many of them due to puerperal fever.

To better understand the situation, we will add a new column to the dataset with the proportion of deaths per year in each clinic.

```{r}
data_death <- data_death %>% 
  mutate(proportion=death*100/births)

```

Compare a situação entre as duas clínicas usando estatísticas sumárias e/ou gráficos que achar relevantes. Explique exatamente o que você está comparando (qual a variável utilizada?).

```{r}
ggplot(data_death, aes(y=death, x=clinical)) +
  geom_boxplot(colour = "darkblue") + 
  labs(
    title = "Distribution of Death in two Clinics",
    x = "Clinic",
    y = "Number of Deaths" )

```

As observed in the graph, there is a difference between the medians of Clinic 1 and Clinic 2. Clinic 1 exhibits a higher median than Clinic 2, which suggests that Clinic 1 records a greater number of deaths. Furthermore, the minimum number of deaths in Clinic 1 exceeds the maximum value observed in Clinic 2, providing additional evidence that Clinic 1 has a higher mortality rate than Clinic 2.

## Buscando por uma explicação e propondo uma intervenção

Dr. Semmelweis observed a pattern in these data and became concerned. The only difference between the two clinics was that, in Clinic 1, medical students attended the deliveries, whereas in Clinic 2 midwives attended the women. While the midwives worked exclusively with childbirth, the students also performed autopsies.

Semmelweis began to suspect that something in the autopsy room was influencing the mortality rate. His primary hypothesis was that some form of contaminating agent was being transmitted on the hands of the students, causing puerperal fever. In an attempt to reduce the high mortality rate, he instructed all staff to wash their hands. At the time, however, this measure was highly controversial, since the scientific knowledge of the period did not yet include an understanding of bacteria.

```{r, message=FALSE}
# Reading the monthly deaths data for Clinic 1
month <- read_csv("mortes_mensais.csv")

# Renaming the columns
month <- month %>%
  rename(
    date = data,
    births = nascimentos,
    death = mortes
  )

# Adding the percentage of deaths relative to births
month <- month %>% mutate(proportion = death * 100 / births)

# Printing the first observations of the data set
head(month)
```

## The Effect of Handwashing

The proportion of deaths over time is analysed through the following plot:

```{r, eval=TRUE}
ggplot(month, aes(x = date, y = proportion)) +
  geom_point() +
  labs(
    title = "Monthly Proportion of Deaths in Clinic 1",
    x = "Date",
    y = "Proportion of Deaths (%)"
  )
```

The plot provides an indication of when the order to wash hands was implemented. The change occurred in the summer of 1847.

```{r}
# Date of the start of the order Handwashing
start_handwashing = as.Date('1847-06-01')
```

```{r}
# Creating a binary column indicating whether the date falls on or after
# the handwashing order (FALSE = before the order, TRUE = on or after)

month <- month %>%
  mutate(order_handwash = date >= start_handwashing)

```

# Let's plot a new line graph to better compare the monthly proportion of the data before and after the handwashing order

```{r,Line and Point Graph, eval=TRUE}

#Line Plot 

ggplot(month, aes(x = date, y = proportion, group = order_handwash, colour = order_handwash)) +
  geom_line() +
  geom_point() +
  # Changing the labels in the legend
  scale_color_discrete(labels = c("FALSE" = "Before", "TRUE" = "After")) +
  labs(
    title = "Monthly Proportion of Deaths in Clinic 1",
    x = "Date",
    y = "Proportion of Deaths (%)",
    colour = "Handwashing Order in Effect"
  )
```

```{r}
# Number of months in which the handwashing order was not yet in effect
count_months_no_order <- function(x) {
  sum(x == FALSE)
}
months_no_order <- count_months_no_order(month$order_handwash)
```

```{r}
# Mean proportion of deaths in the period before the handwashing order
mean_proportion_death_no_handwash <- mean(month$proportion[month$order_handwash == FALSE])
mean_proportion_death_no_handwash
```

```{r}
# Number of months in which the handwashing order was in effect
months_yes_order <- as.numeric(sum(month$order_handwash))
months_yes_order
```

```{r}
# Mean proportion of deaths in the period after the handwashing order
mean_proportion_death_yes_handwash <- mean(month$proportion[month$order_handwash == TRUE])
mean_proportion_death_yes_handwash
```

Comparing the situation, focusing on the proportion of death between wahs or not wahse order. Compare a situação, com relação à proporção de mortes, entre as duas situações (lavar ou não as mãos) usando um boxplot e estatísticas sumárias que achar relevantes.

```{r, Box Plot Graph}

ggplot(month, aes(y=proportion, x=order_handwash, color=order_handwash))+ 
  geom_boxplot() + 
  # Changing the labels on the x-axis
  scale_x_discrete(labels = c("FALSE" = "Before", "TRUE" = "After")) +
  # Changing the labels in the legend
  scale_color_discrete(labels = c("FALSE" = "Before", "TRUE" = "After")) +
  labs(
    title = "Distribution of Death Monthly in Clinic 1 ",
    x = "Date",
    y = "Proportion of Deaths (%)",
    colour = "Handwashing Order in Effect"
  )

```

```{r}
# Calculating the median of the proportion of deaths

median_proportion_death_no_handwash <- median(month$proportion[month$order_handwash == FALSE])
median_proportion_death_yes_handwash <- median(month$proportion[month$order_handwash == TRUE])
```

```{r}

#Calculating the mode of the distribution

# Function to calculate the mode using Czuber's formula
# Mo = L + [d1 / (d1 + d2)] * h
modal_czuber <- function(x) {
  x <- na.omit(x)                              # remove NA
  h <- hist(x, plot = FALSE)                   # create the classes (bins)
  counts <- h$counts
  i <- which.max(counts)                       # modal class position

  L <- h$breaks[i]                             # lower limit of the modal class
  width <- h$breaks[i + 1] - h$breaks[i]       # class width (h)

  f_prev <- if (i == 1) 0 else counts[i - 1]                  # previous class frequency
  f_next <- if (i == length(counts)) 0 else counts[i + 1]     # next class frequency

  d1 <- counts[i] - f_prev                     # modal minus previous
  d2 <- counts[i] - f_next                     # modal minus next

  L + (d1 / (d1 + d2)) * width                 # Czuber's formula
}

# Mode with handwashing
mode_proportion_death_yes_handwash <- modal_czuber(month$proportion[month$order_handwash == TRUE])

# Mode without handwashing
mode_proportion_death_no_handwash <- modal_czuber(month$proportion[month$order_handwash == FALSE])

```

```{r, echo=FALSE}
# Difference between mean 

difference_mean <- mean_proportion_death_no_handwash - mean_proportion_death_yes_handwash

difference_median <- median_proportion_death_no_handwash - median_proportion_death_yes_handwash

```

According to the statistical summary of the monthly proportion of deaths in Clinic 1, the period before the handwashing order had a *mean* =`r mean_proportion_death_no_handwash`, a *median* = `r median_proportion_death_no_handwash`, and a *mode* = `r mode_proportion_death_no_handwash`. In the period after the order, the *mean* was `r mean_proportion_death_yes_handwash`, the *median* was `r median_proportion_death_yes_handwash`, and the *mode* was `r mode_proportion_death_yes_handwash`. Although the modes of the two distributions are similar, the mean differs by `r difference_mean` percentage points and the median by `r difference_median` percentage points. Both measures differ by approximately 8 percentage points between the periods. Furthermore, the boxplot shows that the distribution before the order was highly dispersed and positively skewed.

## Testing Hyphotesis

Despite the marked decrease in the proportion of deaths observed after the students began washing their hands, Dr. Semmelweis was met with disbelief. At the time, he did not have the statistical tools needed to defend his position. In this study, we use a hypothesis test to draw conclusions from the data.

The mean proportion of deaths differs between the two periods (before and after the handwashing order). Can this difference be attributed to handwashing, or could it plausibly arise by chance alone, even if handwashing had no practical effect? To address this question, we adopt as the null hypothesis (H~0~) that handwashing has no effect on the proportion of deaths, and we assess how likely a difference as large as the one observed would be under that assumption. If this probability is very small, we have evidence against H~0~ and in favor of an effect of handwashing.
